An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: A Call for Human-Centered Higher Education

### 2024-07-02 paper/paper.qmd
Paper submitted to Computers & Education

---
title: "An AI is Haunting Campus"
format: docx
bibliography:
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

## Background

> _Mission Statement: Our organization seeks to level up on all matters related._  
> -- An Organization Run by its Marketing Department

On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language. The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels: by early December, ChatGPT was being panned as "the best artificial intelligence chatbot ever released to the general public" [@roose_brilliance_2023], and a tool able to generate "impressively detailed" and "human-like" text, leading some to wonder whether it could "replace humans" [@lock_what_2022].

Universities, much like every other organization dealing with information, had to respond in a compressed span of time to the rapid release of generative AI (GenAI) tools like ChatGPT, most of them created and unleashed in the wild by a handful of large American corporations including [Meta](https://en.wikipedia.org/wiki/Meta_Platforms) (or were we to deadname it, facebook), [OpenAI](https://en.wikipedia.org/wiki/OpenAI), [Google](https://en.wikipedia.org/wiki/Alphabet_Inc.), and [Microsoft](https://en.wikipedia.org/wiki/Microsoft).

GenAI was hailed as nothing short of revolutionary. McMaster, for example, [stated](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) that:

> [u]nderstanding how generative AI works, and how it can support work within organizations represents an ongoing area of exploration and innovation. The opportunities of generative AI are exciting: creating workplace efficiencies to allow for different kinds of work, adding capabilities for individuals and teams, and offering personalized uses.

AI was quickly painted as an _innovative_ and _exciting_ technology, that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. It seemed inevitable that universities _had to do something_ with it, at the risk of being let behind.

The fear of missing out, combined with a modicum of concern for the ethical implications of GenAIs^[Ethical concerns for the most part have had to do with policing academic dishonesty among students.], has led to institutional responses that have been less than coherent. A case in point is Emory University, where the university gave a $10,000 prize to a group of students who proposed developing an AI tool for an Entrepreneurship Summit, only for the students to be subsequently suspended by the school's Honor Council for the potential of their tool for cheating^[There was no evidence that cheating had happened.] [@koebler__university_2024].

For its part, McMaster was a leader in responding to the rapid, and disruptive, release of these _revolutionary_ tools, and its provisional [guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/) (released under a Creative Commons Licence) have since inspired those of other post-secondary institutions in Canada, including the University of Guelph [-@office_of_teaching_and_learning_provisional_2023] and Trent University [-@trent_university_generative_2024].

The aim of the provisional guidelines was to inform the university community about the use of AI in "operational excellence". As we approach the Fall Term of 2024, with almost two years of cohabitation experience with AI for the masses at our backs, the time has come to update these guidelines. Pertinently, the institution, via its Office of the Provost, [launched a committee](#ai-committee-announced) to explore the use of AI across campus. This committee is composed of three sub-committees tasked with examining the use of AI in 1) teaching and learning, 2) research, and 3) operational excellence in "our work".

The sub-committees, in turn, convened panels to [provide recommendations](#ai-panel-tasks) on updates to the [Provisional Guidelines](https://provost.mcmaster.ca/office-of-the-provost-2/generative-artificial-intelligence-2/generative-ai-for-staff/provisional-guidelines-on-the-use-of-generative-ai-in-operational-excellence/). The [priority initiatives](#ai-panel-tasks) for the members of the AI Teaching and Learning Panel are to help shape the guidelines for teaching and learning at McMaster, a consequential task "for how educators take up generative AI in their course and assessment design and how students approach the use of this technology".

According to the announcement of this AI Committee at McMaster:

> AI is a powerful tool and I believe that universities have an important role to play in its adoption.  
> -- Susan Tighe, Provost

Notice that this statement (like the definition of the work before the panels above), takes for granted that 1) AIs are powerful tools, and 2) universities can be leading players in their adoption. This attitude is almost fatalistic^[In the sense of "fated to".] in its enthusiasm, and perfectly consistent with the hype observed around AI this time^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent of several cycles of hype [see @siegel_ai_2023]--fundamentally a vastly overblown tech bubble [@vinsel_dont_2023].] [@angwin_press_2024].

To put things in perspective, hammers are powerful tools. Word processors are powerful tools. Lasers are powerful tools. And yet we do not have high-level university committees for each and every tool that is used at the institution. What is it about AIs that, unlike many other tools, requires a pan-university consultation on an "_ethical approach to shaping policies and developing new ways of using AI to ensure a responsible and beneficial integration where it makes sense_"?

(The last part of the statement above carries quite a bit of weight here: where does adoption of AI make sense? Presumably this is what the Committee on AI will try to answer.)

To understand why many organizations, including McMaster University, have decided to spend considerable resources studying the adoption and responsible use of AIs, it is essential to understand what the technology claims to do that sets it apart from other tools, and what it in actuality does.

In doing this, it is essential that we remain alert to the fact that technologies are _never_ politically neutral, and therefore it is important to ask two questions when considering their adoption:

@@

We also need to understand the goals of the organization that considers using the technology. When does it make sense for a university like McMaster to "integrate" AIs as part of its core mission, accounting for what the technology does for some people and what it does to others?

Let us begin with the mission of McMaster as an institution of higher education.

@@

McMaster's [mission](https://president.mcmaster.ca/mission-vision/) is "the discovery, communication and preservation of knowledge," accomplished with a commitment to "creativity, innovation and excellence". These aspirations must respond to some underlying values, and indeed, the university explicitly values "integrity, quality, inclusiveness and teamwork". Ultimately, the university aims to "inspire critical thinking, personal growth, and a passion for lifelong learning" and to serve "the social, cultural, and economic needs of our community and our society."

A mission statement is like an onion, and one needs to peel multiple layers to try to understand the core principles enunciated there. Why are things like "integrity" and "critical thought" valued? Or, what happens when "integrity", "quality", and/or "excellence" clash with "the economic needs of our society" (say, a mandate by democratically elected governments to spend as little as possible in post-secondary education^[Between 2000-2020, Ontario had the lowest spending on post-secondary education of all provinces in Canada, relative to its share of spending [@caut_provincial]. Public funding of universities in Ontario as percent of revenue fell from 45.6% in 2001/2 to 35.5% in 2019/20 [@statistics_canada_college_2022]. As noted by @usher_state_2023, Ontario is "a province where underfunding [higher-education] institutions is a pan-partisan affair". Is it any wonder that universities in the province are pushed to find "efficiencies"?])?

Clearly, there is some tension implicit in the mission of a university. But being generous, we can take the mission statement at face value and assume that the highest priorities are those related to knowledge obtained, preserved, and communicated with creativity, integrity, excellence, and quality.

And so, what do AIs do that can advance the mission of the university?

## What do AIs do?{#what-do-ais-do}

@@

A key reason why many organizations have decided to spend substantial resources considering the adoption of AIs is that, unlike most other tools, they have been presented quite deliberately as _everything tools_^[Almost as flexible as, well, a naturally intelligent human, but faster and supposedly cheaper. [Here](https://www.youtube.com/watch?v=yUoj9B8OpR8) is OpenAi's Mira Murati talking about "AI Everywhere: Transforming Our World, Empowering Humanity"--just maybe not all of humanity: "some creative jobs maybe will go away, but maybe they shouldn't have been there in the first place". One can be sure she means the jobs that produced the materials that OpenAI stole to train its AIs.]. While hammers are understood to have a limited range of uses that mostly have to do with concentrating force, and word processors are generally only used to record words, the release of GenAIs to the masses was accompanied by innumerable lofty promises^[Quite a few of them made by entities and individuals that had a stake in the adoption of the technology, ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].

@@

Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" by "automating customer self-service, accelerating response times on an increasing number of tasks as well as providing greater accuracy, enhanced routing and intelligent context gathering" [@ibm_what_2023].

According to Allerin^[Curiously, Allerin is not identified as an artificial intelligence company in @joshi_5_2022 but only in the much later article by @garsten_what_2024.], AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024]. But why stop at self-driving vehicles? McMaster University's Canada Excellence Research Chair Laureate Ali Emadi "hopes to help revolutionize and influence everything from personal travel to city planning" by putting AI directly into our transportation systems [@dorey_driving_2023].

In mental health care, chatbots could complement human therapists "via comprehensive data access and [analysis of] behavioral patterns" to "mimic practitioner questions and subsequently make recommendations based on a user’s inputs" [@silva_4_2023].

AIs will even write code, perhaps helping themselves to some self-programming along the way: AIs that write themselves, or AI self-help?

```{r ai-writing-code, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}An example of Atlassian AI writing code"}
knitr::include_graphics(glue::glue(here::here(), "/images/ai-bug-fix.png"))
```

The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves, all the way down to micromanaging intimate facets of individual lives, for example by revolutionizing dating through the use of chatbots to help "people connect and find potential partners" (who presumably also use chatbots to do the same) [@khalatian_matchmaking_2023]. Not even the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html) are beyond the revolutionary all-seeing eye of AI.

```{r ai-litter-box, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}Revolutionizing pet health"}
knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
```

The forecast that AI will improve everything (from global climate to the dumps of our cats) was certain to, sooner or later, reach higher-education, a target-rich environment for selling hype--and indeed, colleges and universities have not been spared the onslaught of promises.

Examples of this abound.

@power_3_2024,^[Rhett Power is a motivational speaker and executive coach.], writing for Forbes, presents three innovative use cases for AI in higher education. First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably to work on something else, and not for the university, hence leading to "efficiencies".]. AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants. AI could also personalize the experience of consumers (occasionally also called students) by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.

Mr. Power's message to the higher-education sector? It is time for institutions to "level up": jump on the bandwagon, fellas, don't let it leave you behind!

In another instance, @hie_how_2023^[Anthony Hié is Chief Innovation and Digital officer at [Excelia](https://www.excelia-group.com/about-excelia/our-schools/excelia-business-school), a top business school in France. Claire Thouary is the founder of [QACE UP](http://qace-up.fr), a consulting firm that supports higher-education institutions in their quest for "quality excellence", whatever that means.] argue that AIs will likely be the primary way to access information in the future. For this reason, institutions must equip students with the skills needed to use AI competently, in particular train them to perfect the art of _prompt engineering_ (the ability to craft effective queries that prompt a GenAI to provide the most useful outputs). Universities will use AIs to improve learning and challenge the critical thinking of students. Faculty will also benefit from AI: it will be used to evaluate their students^[A task of dubious legality, and clearly a way to cheat students of a learning experience.], and will handle repetitive tasks like writing course syllabi^[In my 20+ years of experience as an educator, writing syllabi has been an activity essential to understand my own courses, what I teach and how I teach it; automating the task of writing syllabi is another way of degrading the quality of instruction that students can expect to receive.].

@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). It did not take long for this prediction to turn sourer than beans in August: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it.  Shortly after that, Meta, the main cheerleader of the eponymous metaverse, [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.

\newpage

Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape, including by @johnson_ai_2023^[[Gridserve](https://www.gridserve.co.uk/about-us/) is a UK company "dedicated to enhancing the technological capabilities of the education sector through customised information and communication technology (ICT) services".] and @contact_north_five_2024^[teachonline.ca by [Contact North/Contact Nord](https://teachonline.ca), is "a resource for post-secondary faculty and instructors...to find the latest information on new technology and new developments in online learning, as well as practical tools and resources to help them integrate technology in their teaching in a way that improves the learning experience for their students". Contact North/Contact Nord is a not-for-profit corporation funded by the Government of Ontario, the province that has underfunded its post-secondary education sector for decades [@usher_state_2023].]. A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original, not "could", or "maybe will"):

@@

A benefit of portraying AI as an _everything tool_ (for the proponents of the technology, at least), is that conversations around it quickly turn into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. To be sure, AIs can do _some_ things well, but critics are forced to respond to every single propounded use case with in-depth analysis, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself^[This communication strategy is what Trump's erstwhile and possibly future advisor Steve Bannon termed ["flooding the zone with shit"](https://www.vox.com/policy-and-politics/2020/1/16/20991816/impeachment-trial-trump-bannon-misinformation). Perhaps not coincidentally, AI is now a major vector of disinformation, flooding the zone at a rate that humans cannot hope to counteract [@maiberg__google_2024].]. Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do do?

A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.

In what follows, it is not my intention to confront every single possible use case of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop^[I have already spent more than 90 hours working on this essay, and it is still nowhere in the vicinity of being an exhaustive apology or indictment of AI.], namely to defeat criticism by flooding the zone. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the moral, environmental, and social compromises demanded by AI^[Most discussions of AI as an everything tool acknowledge that the use of AIs is not without problems. For instance, @mollick_instructors_2024 note that "Large Language Models are trained in ways that may violate copyright, and often rely on the efforts of low-wage workers in precarious conditions for part of their training. Models are trained on biased data and can produce either subtly or overtly biased results. And because these biases seem to come from an objective machine, they can be especially pernicious. Using AI systems can mean sharing private data with the for-profit companies developing LLMs, and that data may be used to train further AIs." Therefore, "educators may want to consider before deciding whether to use these systems". This disclaimer, consuming fully one third of a page in a 76-page document dedicated to extolling an AI-enabled future in education, highlights the important role of institutions to provide the moral and ethical clarity that may elude individual instructors working under desperately under-resourced conditions, or simply willing to win a race to the bottom.].

## But really, what do AIs do? {#really-what-do-they-do}

GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning many, many, _many_ hidden layers. As an example, Large Language Models (such as those behind OpenAI's ChatGPT, Microsoft's Copilot, Google's Gemini, and facebook's LLaMA) are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

But, what are [neural](https://en.wikipedia.org/wiki/Large_language_model#Further_reading) [networks](https://medium.com/@asifurrahmanaust/lesson-3-neural-network-is-nothing-but-a-linear-regression-e05a328a0f23)?

Behind all the jargon, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021]. In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure, but regression models nonetheless [@ripley_neural_1994]. Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing an input (e.g., a prompt) to a "statistically probable" outcome.

@@

Below is an example of the simplest regression model with only one independent variable (i.e., a bivariate model). The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$. This model needs only two parameters: an intercept (the conditianl mean $\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$. Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.
```{r simulate-simple-example}
set.seed(13253)

@@

```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Simple Example of a Regression Model"}
ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="70%", fig.cap="Training data and model output compared"}

@@

Neural networks do something similar but using a net-like architecture that is more flexible at identifying the characteristics of the data. For instance, the following figure shows a relatively simple neural network trained using using the same data. This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers. In comparison to the linear regression and its two parameters, this neural network uses seven parameters. Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contingent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

```{r plot-shallow-neural-network, out.width="70%", fig.cap="Simple Example of a Shallow Neural Network"}
knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
```

@@

The plot below illustrates the predictions of this neural network, where we can see that the model is not limited to a linear response. Neural networks are more flexible than linear regression, and they achieve this by being less parsimonious (they use more parameters).

```{r data-vs-predictions-comparison-2, out.width="70%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
ggplot() + 
  geom_point(data = df_pred_all,
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

More complex neural network can be designed. The following plot is of a network deeper than the previous one, with two hidden layers, each consisting of five neurons. This model uses forty-six parameters. 
```{r train-deeper-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(242)
nn2 <- neuralnet(f, 
                data = df, 
                hidden = c(5, 5), 
                linear.output = T)
```

@@

```{r plot-deeper-neural-network, out.width="70%", fig.cap="Example of a Deeper Neural Network"}
knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
```

@@

The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data.

```{r data-vs-predictions-comparison-3, out.width="70%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

>>>>

---
title: "An AI is Haunting Campus^[Karl Marx wrote about a \"spectre...haunting Europe--the spectre of communism\"; GenAI looks awfully like the spectre of unregulated surveillance capitalism.]: A Call for Human-Centered Higher Education"
format: docx
bibliography:
  - "`r system('kpsewhich ../bibliography/bibliography.bib', intern=TRUE)`"
---

@@

## Introduction

On November 30, 2022, OpenAI released what became its flagship product, [ChatGPT](https://en.wikipedia.org/wiki/ChatGPT), a chatbot based on [Large Language Models](https://en.wikipedia.org/wiki/Large_language_model) (LLMs) designed and trained to mimic natural language. The launch of the chatbot was an impressive feat of marketing [@cowen_chatgpt_2023], and it took less than two weeks for the hype to reach stratospheric levels [@roose_brilliance_2023; @lock_what_2022]. Soon after, seemingly everyone wanted in on the Generative AI (GenAI) craze^[The pattern of cyclical hypes around AI even has a name: [AI winters](https://en.wikipedia.org/wiki/AI_winter); the current one, with chatbots playing the loudest horn atop the bandwagon, is just the most recent cycle of hype [see @siegel_ai_2023].], and no surprise, given how the technology was hailed as nothing short of revolutionary.

According to some, AI technologies in education "offer new tools and applications that have the potential to transform traditional teaching and learning methods" [@adiguzel_revolutionizing_2023, p. 1], and have "the potential to fundamentally change the classroom experience and the knowledge and skills outcomes of our students" [@dwivedi_opinion_2023, p.20]. Furthermore, the technology "is here to stay...[t]herefore, teachers and students must develop the specific digital competencies needed to use such tools in ways that are pedagogically beneficial and ethical" [@kohnke_chatgpt_2023, p. 546].

AI was quickly painted as an _innovative_ and _exciting_ technology, capable of _enhancing_ research and education, and that moreover can lead to _efficiencies_^["Efficiencies", to be clear, is the day's euphemism for cutting costs [@oconnell_let_2012, p. 64]. Cost-cutting is usually implemented by finding "redundancies" rather than finding better things for people to do.]. Institutions of higher education were compelled to get in on the act, forming high-level advisory committees and drafting guidelines to guide the university community leverage AI in principled ways [e.g., @office_of_teaching_and_learning_provisional_2023; @mcmaster_university_provisional_2024; @trent_university_generative_2024; @university_of_toronto_ai_2024].

For perspective, hammers are powerful tools, and so are thousands of other instruments found at universities. And yet not every single one commands a high-level committee to advice on its use. Just what is it about AIs that requires pan-university consultations on how to use them? To understand why practically every institution of higher education has decided to spend considerable resources studying the adoption and responsible use of AI, one needs to understand what the technology claims to do that sets it apart from other tools--and then, examine what it actually does.

In doing this, we must remain alert to the fact that technologies are never politically neutral, and two questions must be asked when considering their adoption:

@@

As well, we need to understand the character of the organization that considers the use of the technology.

This essay aims to provide some clarification around 1) the things that GenAIs claim to do in education; and 2) what the technology actually does; within 3) the context of a principled organization. I conclude that the promise of generative AI has been overstated, that is is unlikely to deliver, and that in any case the technology, as sold by the tech industry, is antithetical to many of the values that universities claim to hold dear.

@@

Universities present themselves as institutions that serve "the social, cultural, and economic needs of [their] community and [their] society" [@office_of_the_president_mcmaster_university_mission_2024]; that strive "to make valued and socially responsible contributions to our local communities, to Canada, and to the world" [@trent_university_vision_2024]; and that "serve society and...enhance the quality of life through scholarship" [@university_of_guelph_mission_2024].

This overarching goal is achieved through "the discovery, communication and preservation of knowledge" [@office_of_the_president_mcmaster_university_mission_2024]; a dedication "to fostering an academic community in which the learning and scholarship of every member may flourish" [@the_office_of_the_governing_council_secretariat_institutional_1992]; by "[encouraging] and [celebrating] excellence and innovation in teaching, learning, research and student development" [@trent_university_vision_2024]; with a determination "to put the learner at the centre of all [the university] does" [@university_of_guelph_mission_2024].

Unprincipled organizations do not need to question what the adoption of a technology does to and for whom--as long as it advances whatever its goal is at any particular moment. But universities aim to achieve their mission while adhering to "integrity, quality, inclusiveness and teamwork" [@office_of_the_president_mcmaster_university_mission_2024]; with "vigilant protection for individual human rights, and a resolute commitment to the principles of equal opportunity, equity and justice" [@the_office_of_the_governing_council_secretariat_institutional_1992]; and  as they "foster sustainability, in its environmental, social and economic dimensions" [@university_of_guelph_mission_2024].

Universities must strike a careful balance between what they aim to achieve and how they achieve it. And so, what do AIs do that can advance the mission of universities without compromising their core values?

## What do GenAIs do?{#what-do-ais-do}

@@

A reason why many organizations have spent substantial resources considering the adoption of AIs is that they have been presented, quite deliberately, as _everything tools_: tools capable of doing many things a human can do, but faster and supposedly cheaper. While hammers are understood to have a limited range of uses that mostly have to do with blunt force, the release of GenAIs to the masses was accompanied by innumerable promises--quite a few of them made by entities and individuals that had a stake in the adoption of the technology^[Ranging from "AI is inevitable, and I will sell it to you" to "AI is inevitable, and for a consulting fee I will tell you how to level up with it".].

@@

Also, that LLMs would transform sectors as diverse as health care, finance, human resources, insurance, "and beyond" [@ibm_what_2023]. AI-powered self-driving cars would reduce accidents, car ownership, pollution and noise, and would automate logistics and make parking tickets obsolete [@joshi_5_2022; @garsten_what_2024]. In mental health care, chatbots could complement human therapists by using patient data to analyze behavioral patterns to then mimic practitioner questions [@silva_4_2023]. AIs would revolutionize dating by helping "people connect and find potential partners" [@khalatian_matchmaking_2023].

The list of promises about what AI _can_ do has grown to encompass pretty much everything, from saving the planet from ourselves to micromanaging the [bowel movements of our pets](https://finance.yahoo.com/news/next-generation-ai-powered-cat-140000795.html).

```{r ai-litter-box, out.width="70%", fig.cap="\\label{fig:ai-bug-fix}Revolutionizing pet health"}
#| label: fig-revolutionizing-pet-care

knitr::include_graphics(glue::glue(here::here(), "/images/ai-litter-box.png"))
```

Colleges and universities have not been spared the onslaught of promises.

@power_3_2024, presents three innovative uses for AI in higher education. First, AI could take over manual tasks and improve the budget by releasing administrative staff^[Probably from working at the university]. AI could evaluate applications for admissions, including assessing traits like the "grit" and "empathy" of applicants. AI could also personalize the experience of consumers by predicting their final grade with 60-70\% accuracy, and by providing corrective and instructive feedback.

In another instance, @hie_how_2023 argue that AIs will likely be the primary way to access information in the future. For this reason, institutions must equip students with the skills like _prompt engineering_, the ability to craft effective queries that prompt a GenAI to provide the most useful outputs. Universities will use AIs to improve learning and challenge the critical thinking of students. Faculty will also benefit from AI: it will be used to evaluate their students, and will handle repetitive tasks like writing course syllabi.

@hie_how_2023 warn that "AI is here to stay"^[Writing in 2022, Mr. Hié predicted that the metaverse would be a [revolution in higher education](https://www.linkedin.com/pulse/metaverse-dedicated-pedagogy-revolution-higher-education-anthony-hi%C3%A9). This prediction aged poorly: on February, 2023, Microsoft [disbanded](https://www.theinformation.com/articles/microsoft-kills-its-industrial-metaverse-team-after-4-months) its Industrial Metaverse Core Team scantly four months after launching it. Shortly after that, Meta [slashed its own metaverse team](https://www.reuters.com/technology/meta-lay-off-employees-metaverse-silicon-unit-wednesday-2023-10-03/) in October, 2023.], and that "[faculty] might not be able to stop the AI trend from growing" so they might as well get on with the program and try to shape it as best they can.

Similarly hyperbolic claims about AI in higher education can be found peppered throughout the landscape [e.g., @adiguzel_revolutionizing_2023; @contact_north_five_2024; @dwivedi_opinion_2023; @johnson_ai_2023; ]. A fairly comprehensive list [from @johnson_ai_2023] is as follows (notice the use of "can" in the original):

@@

Discussions of AI as an _everything tool_ quickly devolve into a [Gish Gallop](https://en.wikipedia.org/wiki/Gish_gallop), a rhetorical device where a party makes an overwhelming number of arguments with little regard to their accuracy or strength. To be sure, AIs can do _some_ things well, but critics are forced to respond to every single propounded use with evidence, which usually is much more time consuming than making the unsubstantiated and/or misleading claims of the gallop itself. Proponents of AI can always admit that, ok, AIs do not actually do _that one thing_, but what about the myriad of other things that people say they really do? A second discursive device deployed by proponents of AI is an informal fallacy known as [The False Dilemma](https://en.wikipedia.org/wiki/False_dilemma): adopt AI or fail to level up.

In what follows, it is not my intention to confront every single possible use of AI in teaching and learning, since doing so plays into the intent of the Gish Gallop. As well, I reject the dilemma posed by the purveyors of hype, and I remain convinced that there are multiple ways to excel at teaching and learning that do not require AI, let alone the compromises demanded by AI.

## But what do GenAIs really do? {#really-what-do-they-do}

GenAIs are _deep learning models_ [@ibm_what_2021], so-called because they are neural networks with _depth_, meaning _many_ hidden layers. As an example, Large Language Models are models built with multiple layers of neural networks trained on _very_ large amounts of textual data [@ibm_what_2023].

Behind all the jargon, though, the neural networks that underpin GenAI are models designed to produce "statistically probable outputs" when prompted [@ibm_what_2021]. In more conventional terms, neural networks are regression models--much more sophisticated and powerful than linear regression to be sure--but regression models nonetheless [@ripley_neural_1994]. Here, it is important to note that the term regression refers to [regression to the mean](https://en.wikipedia.org/wiki/Regression_toward_the_mean)--regressing an input (e.g., a prompt) to a "statistically probable" outcome.

@@

@fig-bivariate-model is an example of the simplest regression model with only one independent variable (i.e., a bivariate model). The plot shows the pairs of $y$ and $x$ values, and the line is the model, i.e., the conditional mean $\hat{y}$. This model needs only two parameters: an intercept (the conditianl mean $\hat{y}$ when $x = 0$), and the slope of the line, in other words, the rate of change of $\hat{y}$ with respect to $x$. Each additional independent variable uses one additional parameter to describe the slope of the model with respect to that variable.

```{r simulate-simple-example}
set.seed(13253)

@@

```{r plot-simulated-data, message=FALSE, out.width="70%", fig.cap="Bivariate Regression Model"}
#| label: fig-bivariate-model

ggplot(df,
       aes(x = x,
           y = y)) +
  geom_point() +
  geom_smooth(method = "lm") + 
  theme_minimal()
```

```{r data-vs-predictions-comparison, out.width="70%", fig.cap="Training data and model output compared"}
#| label: fig-data-vs-preds-01

@@

Neural networks do something similar but using a more flexible net-like architecture. For instance, @fig-nn1 shows a relatively simple neural network trained using using the same data. This network is not "deep", as it has only one "hidden layer" comprised of two neurons between the input and the output layers. In comparison to the linear regression and its two parameters, this neural network uses seven parameters. Another key difference between a linear regression model and a neural network is that the latter requires an initial set of parameters, which are often drawn at random, thus making the results of the model contingent on the starting conditions of the training process.
```{r}
# Given the dataframe grid.1, let us now train a neural network. We need to first define a formula that specifies the dependent variable and the independent variables.
f <- as.formula("y ~ x")
```

@@

```{r plot-shallow-neural-network, out.width="70%", fig.cap="Simple Example of a Shallow Neural Network"}
#| label: fig-nn1

knitr::include_graphics(glue::glue(here::here(), "/images/nn1.png"))
```

@@

@fig-data-vs-preds-02 illustrates the predictions of the neural network in @fig-nn1, where we can see that the model is not limited to a linear response. This flexibility by the neural network is achieved by being less parsimonious (it uses more parameters).

```{r data-vs-predictions-comparison-2, out.width="70%", fig.cap="Training data and model outputs compared: linear regression and shallow neural network"}
#| label: fig-data-vs-preds-02

More complex neural networks can be designed. @fig-nn2 shows a network deeper than the previous one, with two hidden layers, each consisting of five neurons. This model uses forty-six parameters. 
```{r train-deeper-neural-network}
# Train a neural network using only two neurons in one hidden layer. The argument `linear.output` is set to TRUE, so that the activation function is not applied to the output node. The default activation function is the logistic function.
set.seed(242)
nn2 <- neuralnet(f, 
                data = df, 
                hidden = c(5, 5), 
                linear.output = T)
```

@@

```{r plot-deeper-neural-network, out.width="70%", fig.cap="Example of a Deeper Neural Network"}
#| label: fig-nn2

knitr::include_graphics(glue::glue(here::here(), "/images/nn2.png"))
```

@@

The increased complexity leads to greater flexibility, and we can see that the model's predictions resemble more closely the actual data (@fig-data-vs-preds-03).

```{r data-vs-predictions-comparison-3, out.width="70%", fig.cap="Training data and model outputs compared: shallow neural network and deeper neural network"}
#| label: fig-data-vs-preds-03

ggplot() + 
  geom_point(data = df_pred_all |>
               filter(type != "regression output"),
  aes(x = x,
      y = y,
      color = type,
      shape = type,
      size = type)) +
  scale_color_manual(values = c("training data" = "black",
                                "regression output" = "blue",
                                "nn1 output" = "chocolate1",
                                "nn2 output" = "brown")) +
  scale_shape_manual(values = c("training data" = 16,
                                "regression output" = 17,
                                "nn1 output" = 18,
                                "nn2 output" = 19)) +
  scale_size_manual(values = c("training data" = 1,
                                "regression output" = 2,
                                "nn1 output" = 2,
                                "nn2 output" = 2)) +
  theme_minimal() +
  theme(legend.position = "bottom")
```
